---
title: "Class 12"
author: "Cienna Santos (PID: A17581026)"
format: pdf
---

# 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble <https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39805981-39949988;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel>

here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population.

Let's now dig into this further.

##  Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Let's make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```

